---
title: "Plastic Waste Dashboard"
author: "Tyler Kinney, Greyson Ballesteros"
format:
  dashboard:
    theme: journal
    orientation: rows
    scrolling: true
    logo: plastic-logo.png
    pages: true
    nav-buttons:
      - icon: github
        href: https://github.com/Tyler-Kinney/STAT-431-Project-1
        aria-label: GitHub Repository
      - icon: envelope
        href: mailto:kinneytyler101@gmail.com
        aria-label: Email
execute:
  echo: false
  error: false
  warning: false
  message: false
editor: source
---

```{r}
#| label: packages-setup
library(tidyverse)
library(glue)
library(here)
library(jsonlite)
library(purrr)
library(ggplot2)
library(ggridges)
library(gt)
library(scales)
library(ggtext)
library(tidytuesdayR)
library(plotly)
library(leaflet)
```

```{r}
#| label: load-plastic-data
# Break Free From Plastic data from TidyTuesday
# https://github.com/rfordatascience/tidytuesday/tree/main/data/2021/2021-01-26
tuesdata <- tidytuesdayR::tt_load('2021-01-26')
plastics <- tuesdata$plastics

journal_colors <- c("#FFFFFF", "#F8D6D6", "#F1ADAD", "#EA8585", "#E35C5C", "#DE4C4C", "#D95C5C")
```
# Plastic Analysis

## Row {height="22%"}

::: {.card}
### Project Purpose

This dashboard explores global plastic waste patterns using the Break Free From Plastic brand audit data from 2019 and 2020.

**RQ1:** Has the composition of plastic types changed from 2019 to 2020?  
**RQ2:** How does plastic waste differ across country income levels?
:::

::: {.card}
### Data Sources

The dashboard combines Break Free From Plastic data from TidyTuesday with GDP data, REST Countries API country-level information, and country code data used to connect countries to API records.
:::

## Row {height="70%"}

### RQ1: Plastic Type Composition

::: {.panel-tabset}

#### Proportion Summary

```{r}
#| label: rq1-proportion-summary
plastic_types <- c("hdpe", "ldpe", "pet", "pp", "ps", "pvc", "o")

Prop_change <- function(type) {
  plastics |>
    filter(parent_company != "Grand Total") |>
    group_by(year) |>
    summarize(
      total_type = sum(across(all_of(type)), na.rm = TRUE),
      total_all = sum(grand_total, na.rm = TRUE),
      .groups = "drop"
    ) |>
    mutate(prop = total_type / total_all) |>
    arrange(year) |>
    summarize(
      plastic_type = type,
      prop_2019 = prop[year == 2019],
      prop_2020 = prop[year == 2020],
      change = prop_2020 - prop_2019
    )
}

map_dfr(plastic_types, Prop_change) |>
  arrange(change) |>
  gt() |>
  tab_header(title = md("**Proportions of Each Plastic Type**"),
             subtitle = "Change in proportion from 2019 to 2020") |>
  fmt_percent(columns = c(change, prop_2019, prop_2020), decimals = 2) |>
  cols_label(plastic_type = "Type",
             prop_2019 = "2019",
             prop_2020 = "2020",
             change = "Change") |>
  tab_options(heading.background.color = "#EB6864",
              column_labels.background.color = "lightgrey",
              row.striping.background_color = "lightgrey",
              row.striping.include_table_body = TRUE) |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels(everything())) |>
  tab_style(style = cell_borders(sides = "bottom",
                                 color = "#333333",
                                 weight = px(2)),
            locations = list(cells_column_labels(), cells_title(groups = "subtitle"))) |>
  data_color(columns = change,
             colors = col_numeric(
               palette = c("lightgreen", "white", "#D95C5C"),
               domain = NULL))
```

#### Slope Chart

```{r}
#| label: rq1-slope-chart
#| fig-alt: "Slope chart showing changes in proportions of each plastic type used from 2019 to 2020."
#| fig-cap: "Change in plastic type composition from 2019 to 2020."
plastics |>
  pivot_longer(cols = hdpe:pvc,
               names_to = "type",
               values_to = "count") |>
  filter(parent_company != "Grand Total",
         !is.na(count)) |>
  group_by(year, type) |>
  summarize(total = sum(count, na.rm = TRUE), .groups = "drop") |>
  group_by(year) |>
  mutate(prop = total / sum(total)) |>
  ungroup() |>
  select(year, type, prop) |>
  pivot_wider(names_from = year,
              values_from = prop,
              names_prefix = "year_") |>
  mutate(change = year_2020 - year_2019,
         type = reorder(type, change)) |>
  ggplot() +
  geom_segment(aes(x = year_2019, xend = year_2020,
                   y = type, yend = type),
               color = "#333333",
               linewidth = 1) +
  geom_point(aes(x = year_2019, y = type), color = "grey", size = 3) +
  geom_point(aes(x = year_2020, y = type), color = "#EB6864", size = 3) +
  scale_x_continuous(labels = scales::label_percent()) +
  labs(title = "Change in Plastic Type Composition",
       subtitle = "(<span style='color:grey;'>2019</span> - <span style='color:#EB6864;'>2020</span>)",
       x = "Proportion of total plastic",
       y = "Plastic type") +
  theme_light() +
  theme(plot.subtitle = element_markdown())
```

#### Year-Over-Year Summary

```{r}
#| label: rq1-yoy-summary
plastics |>
  filter(parent_company != "Grand Total") |>
  pivot_longer(cols = c(hdpe, ldpe, o, pet, pp, ps, pvc, grand_total),
               names_to = "plastic_type",
               values_to = "amount") |>
  group_by(plastic_type, year) |>
  summarize(total = sum(amount, na.rm = TRUE), .groups = "drop") |>
  arrange(plastic_type, year) |>
  group_by(plastic_type) |>
  mutate(`2019_2020_YoY` = (total - lag(total)) / lag(total)) |>
  filter(!is.na(`2019_2020_YoY`)) |>
  select(-total, -year) |>
  arrange(`2019_2020_YoY`) |>
  ungroup() |>
  gt() |>
  tab_header(title = md("**Year-Over-Year Increase in Plastic Use**"),
             subtitle = "Calculated from 2020 compared with 2019") |>
  fmt_percent(columns = `2019_2020_YoY`, decimals = 2) |>
  cols_label(plastic_type = "Type",
             `2019_2020_YoY` = "YoY Increase") |>
  tab_options(heading.background.color = "#EB6864",
              column_labels.background.color = "lightgrey",
              row.striping.background_color = "lightgrey",
              row.striping.include_table_body = TRUE) |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels(everything())) |>
  tab_style(style = cell_borders(sides = "bottom",
                                 color = "#333333",
                                 weight = px(2)),
            locations = list(cells_column_labels(), cells_title(groups = "subtitle"))) |>
  data_color(columns = `2019_2020_YoY`,
             colors = col_numeric(
               palette = c("lightgreen", "white", "#D95C5C"),
               domain = NULL))
```

#### Composition by Year

```{r}
#| label: rq1-composition-plot
#| fig-alt: "Segmented bar graph displaying the distribution of plastic types in 2019 and 2020."
#| fig-cap: "Proportion of plastic types used per year from 2019 to 2020."
plastics |>
  pivot_longer(cols = c(hdpe, ldpe, o, pet, pp, ps, pvc),
               names_to = "plastic_type",
               values_to = "plastic_amount") |>
  filter(parent_company != "Grand Total",
         !is.na(plastic_amount)) |>
  group_by(year, plastic_type) |>
  summarize(total = sum(plastic_amount, na.rm = TRUE),
            .groups = "drop") |>
  ggplot(aes(x = factor(year), y = total, fill = plastic_type)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = label_percent()) +
  scale_fill_manual(values = journal_colors) +
  labs(title = "Proportion of Plastic Types by Year",
       x = "Year",
       y = "Proportion",
       fill = "Plastic Type") +
  theme_light()
```
#### Plastic Type Output by Year

```{r}
#| label: rq1-dynamic-visualization
#| fig-alt: "Bar chart comparing proportions of plastic types for 2019 and 2020."
#| fig-cap: "Proportion of global plastic types in 2019 and 2020, showing changes in composition over time."
#| fig-height: 6
#| fig-width: 5
global_plot_data <- plastics |>
  filter(parent_company != "Grand Total") |>
  pivot_longer(
    cols = c(hdpe, ldpe, o, pet, pp, ps, pvc),
    names_to = "plastic_type",
    values_to = "plastic_amount"
  ) |>
  group_by(year, plastic_type) |>
  summarize(total = sum(plastic_amount, na.rm = TRUE), .groups = "drop") |>
  group_by(year) |>
  mutate(proportion = total / sum(total)) |>
  ungroup()

global_plot <- global_plot_data |>
  ggplot(aes(
    x = plastic_type,
    y = proportion,
    fill = factor(year),
    text = paste(
      "Year:", year,
      "<br>Plastic type:", plastic_type,
      "<br>Proportion:", scales::percent(proportion, accuracy = 0.1)
    )
  )) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::label_percent()) +
  scale_fill_manual(values = c("#F1ADAD", "#EB6864")) +
  labs(
    title = "Global Plastic Type Composition",
    x = "Plastic Type",
    y = "Proportion",
    fill = "Year"
  ) +
  theme_light() +
  theme(plot.title = element_text(face = "bold"))

ggplotly(global_plot, tooltip = "text")
```
:::

# China Case Study

## Row {height="18%"}

::: {.card}
### Why China?

China was selected as a case of change because its plastic composition shifts substantially from 2019 to 2020, particularly with a large increase in the “other” plastic category and a decline in PET. These changes highlight how plastic waste composition can vary over time within a single country. Additionally, we were interested in investigating how the COVID-19 pandemic may have impacted China specifically and influenced patterns of plastic usage and waste.

:::

## Row {height="82%"}

::: {.panel-tabset}

#### China Composition

```{r}
#| fig-height: 3
#| fig-width: 8

china_data <- plastics |>
  filter(country == "China",
         parent_company != "Grand Total") |>
  pivot_longer(
    cols = c(hdpe, ldpe, o, pet, pp, ps, pvc),
    names_to = "plastic_type",
    values_to = "plastic_amount"
  ) |>
  group_by(year, plastic_type) |>
  summarize(
    total = sum(plastic_amount, na.rm = TRUE),
    .groups = "drop"
  ) |>
  group_by(year) |>
  mutate(proportion = total / sum(total)) |>
  ungroup()

china_data |>
  ggplot(aes(x = plastic_type, y = proportion, fill = factor(year))) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::label_percent()) +
  scale_fill_manual(values = c("#F1ADAD", "#EB6864")) +
  labs(
    title = "Plastic Type Composition in China (2019 vs 2020)",
    x = "Plastic Type",
    y = "Proportion",
    fill = "Year"
  ) +
  theme_light() +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )
```

:::
