---
title: "Plastic During COVID"
author: "Tyler Kinney, Greyson Ballesteros"
format:
  dashboard:
    theme: journal
    orientation: rows
    scrolling: true
    logo: plastic-logo.png
    embed-resources: true
    nav-buttons:
      - icon: github
        href: https://github.com/Tyler-Kinney/STAT-431-Project-1
        aria-label: GitHub Repository
      - icon: envelope
        href: mailto:kinneytyler101@gmail.com
        aria-label: Email
      - icon: linkedin
        href: https://www.linkedin.com/in/tyler-kinney-7208303a2
        aria-label: LinkedIn Profile
execute:
  echo: false
  error: false
  warning: false
  message: false
editor: source
server: shiny
---


```{r}
#| label: packages-setup

library(tidyverse)
library(here)
library(purrr)
library(ggplot2)
library(ggridges)
library(gt)
library(scales)
library(ggtext)
library(ggrepel)
library(gganimate)
library(gifski)
library(leaflet)
library(sf)
library(rnaturalearth)
library(glue)
library(shiny)
```

```{r}
#| label: load-plastic-data
#| context: setup
# Break Free From Plastic data from TidyTuesday
# https://github.com/rfordatascience/tidytuesday/tree/main/data/2021/2021-01-26
tuesdata <- tidytuesdayR::tt_load('2021-01-26')
plastics <- tuesdata$plastics

journal_colors <- c("#FDECEC", "#F8D6D6", "#F1ADAD", "#EA8585", "#E35C5C", "#DE4C4C", "#D95C5C")
```

# Changing Plastic Type
## Row 1 {height=10%}

### Column 1 {.tabset}
#### Intro
The introduction of novel coronavirus became a defining moment in global history when it made its first public appearance in Wuhan, China. The epidemic spread like wildfire, and has altered the course of humanity to this day. As the virus took its toll, humans began exhausting its resources like never before; one of these resources, of course, was plastics. As students that survived through the peak of COVID-19, we were interested in how the usage of plastics changed from 2019 to 2020, a span of years that has become one of the most relevant turning points in our lives.

This led us to ask: Have the types of plastics that are being used changed significantly from 2019 to 2020?

#### Data Background
In order to answer our research question we obtained plastics data from Break Free from Plastic (BFFP) courtesy of Sarah Sauve. The observational unit is one cleaning project. The data contains variables relating to types of plastic, where the plastic comes from, and country. The columns in the data have been assigned certain classes like "double" or "character". The creators of the data set also cleaned the names of the columns using the janitor package. They also took two data sets from 2019 and 2020 and bound the rows together. The audit was completed in October of 2020, implying that the data was collected during that time frame. The data was collected as part of a bigger project to understand the sources of plastic in different cities.

## Row 2

### Column 1 {.tabset}
#### Animated Bar Chart
```{r}
checkboxGroupInput(
  inputId = "plastic_filter",
  label = "Select plastic types to display:",
  choices = c("hdpe", "ldpe", "o", "pet", "pp", "ps", "pvc"),
  selected = c("hdpe", "ldpe", "o", "pet", "pp", "ps", "pvc"),
  inline = TRUE
)

imageOutput("animated_plastic_plot")
```

```{r}
#| context: server
library(tidyverse)
library(gganimate)
library(gifski)
library(shiny)
library(scales)

output$animated_plastic_plot <- renderImage({

  req(input$plastic_filter)

  outfile <- tempfile(fileext = ".gif")

  filtered_plot_data <- plastics |>
    filter(parent_company != "Grand Total") |>
    pivot_longer(
      cols = c(hdpe, ldpe, o, pet, pp, ps, pvc),
      names_to = "plastic_type",
      values_to = "plastic_amount"
    ) |>
    filter(plastic_type %in% input$plastic_filter) |>
    mutate(plastic_type = factor(plastic_type, levels = input$plastic_filter)) |>
    group_by(year, plastic_type) |>
    summarize(total = sum(plastic_amount, na.rm = TRUE), .groups = "drop") |>
    group_by(year) |>
    mutate(proportion = total / sum(total)) |>
    ungroup()

  validate(
    need(nrow(filtered_plot_data) > 0,
         "No data available for selected plastic types.")
  )

plastic_choices <- c("hdpe", "ldpe", "o", "pet", "pp", "ps", "pvc")

animated_plot <- filtered_plot_data |>
  ggplot(aes(
    x = plastic_type,
    y = proportion,
    fill = plastic_type
  )) +
  geom_col(show.legend = FALSE) +
  scale_y_continuous(labels = label_percent()) +
  scale_fill_manual(
    values = setNames(journal_colors, plastic_choices)
  ) +
  labs(
    title = "Plastic Composition in {closest_state}",
    x = "Plastic Type",
    y = "Proportion"
  ) +
  theme_light() +
  transition_states(
    states = year,
    transition_length = 2,
    state_length = 1
  ) +
  ease_aes("cubic-in-out")

  animate(
    animated_plot,
    nframes = 64,
    fps = 8,
    renderer = gifski_renderer(outfile),
    rewind = FALSE
  )

list(
  src = outfile,
  contentType = "image/gif",
  alt = "Animated plastic composition graph")

}, deleteFile = TRUE)
```

#### Lollipop Chart
```{r}
#| label: rq1-slope-chart
#| fig-alt: "Slope chart showing changes in proportions of each plastic type used from 2019 to 2020."
#| fig-cap: "Change in plastic type composition from 2019 to 2020."
plastics |>
  pivot_longer(cols = hdpe:pvc,
               names_to = "type",
               values_to = "count") |>
  filter(parent_company != "Grand Total",
         !is.na(count)) |>
  group_by(year, type) |>
  summarize(total = sum(count, na.rm = TRUE), .groups = "drop") |>
  group_by(year) |>
  mutate(prop = total / sum(total)) |>
  ungroup() |>
  select(year, type, prop) |>
  pivot_wider(names_from = year,
              values_from = prop,
              names_prefix = "year_") |>
  mutate(change = year_2020 - year_2019,
         type = reorder(type, change)) |>
  ggplot() +
  geom_segment(aes(x = year_2019, xend = year_2020,
                   y = type, yend = type),
               color = "#333333",
               linewidth = 1) +
  geom_point(aes(x = year_2019, y = type), color = "grey", size = 3) +
  geom_point(aes(x = year_2020, y = type), color = "#EB6864", size = 3) +
  scale_x_continuous(labels = label_percent()) +
  labs(title = "Change in Plastic Type Composition",
       subtitle = "(<span style='color:grey;'>2019</span> - <span style='color:#EB6864;'>2020</span>)",
       x = "Proportion of total plastic",
       y = "Plastic type") +
  theme_light() +
  theme(plot.subtitle = element_markdown())
```

### Column 2
#### Interpretation and Table
::: {.card}
Both visuals show the same overall pattern: plastic type composition changed noticeably from 2019 to 2020. The animated chart makes the shift easier to see as a before-and-after transition, while the slope chart shows the direction and size of each plastic type's change more directly. The visuals show a clear shift in plastic composition from 2019 to 2020. The **“other”** category decreases substantially (-22.13%), while **polypropylene** (PP) increases the most (+13.62%). Several other plastic types show smaller increases, and PET remains relatively stable. Overall, plastic waste becomes more concentrated in specific types rather than the broad “other” category.

here are the values of the changes seen in both visuals:
```{r}
#| label: rq1-proportion-summary
plastic_types <- c("hdpe", "ldpe", "pet", "pp", "ps", "pvc", "o")

Prop_change <- function(type) {
  plastics |>
    filter(parent_company != "Grand Total") |>
    group_by(year) |>
    summarize(
      total_type = sum(across(all_of(type)), na.rm = TRUE),
      total_all = sum(grand_total, na.rm = TRUE),
      .groups = "drop"
    ) |>
    mutate(prop = total_type / total_all) |>
    arrange(year) |>
    summarize(
      plastic_type = type,
      prop_2019 = prop[year == 2019],
      prop_2020 = prop[year == 2020],
      change = prop_2020 - prop_2019
    )
}

map_dfr(plastic_types, Prop_change) |>
  arrange(change) |>
  gt() |>
  tab_header(title = md("**Proportions of Each Plastic Type**"),
             subtitle = "Change in proportion from 2019 to 2020") |>
  fmt_percent(columns = c(change, prop_2019, prop_2020), decimals = 2) |>
  cols_label(plastic_type = "Type",
             prop_2019 = "2019",
             prop_2020 = "2020",
             change = "Change") |>
  tab_options(heading.background.color = "#EB6864",
              column_labels.background.color = "lightgrey",
              row.striping.background_color = "lightgrey",
              row.striping.include_table_body = TRUE) |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels(everything())) |>
  tab_style(style = cell_borders(sides = "bottom",
                                 color = "#333333",
                                 weight = px(2)),
            locations = list(cells_column_labels(), cells_title(groups = "subtitle"))) |>
  tab_options(
    data_row.padding = px(18),
    heading.padding = px(10)) |>
  data_color(columns = change,
             colors = col_numeric(
               palette = c("lightgreen", "white", "#D95C5C"),
               domain = NULL))
```
:::

## Row 3

### Column 1
#### YoY
:::{.card}

```{r}
#| label: Summary 2 (non-modified)
#| fig-alt: "Table displaying year over year increase in plastic type usage in 2019 and 2020."

plastics |>
  filter(parent_company != "Grand Total") |>
  pivot_longer(cols = c(hdpe, ldpe, o, pet, pp, ps, pvc, grand_total),
               names_to = "plastic_type",
               values_to = "amount") |>
  group_by(plastic_type, year) |>
  summarize(total = sum(amount, na.rm = TRUE), .groups = "drop") |>
  arrange(plastic_type, year) |>
  group_by(plastic_type) |>
  mutate(`2019_2020_YoY` = (total - lag(total)) / lag(total)) |>
  filter(!is.na(`2019_2020_YoY`)) |>
  select(-total, -year) |>
  arrange(`2019_2020_YoY`) |>
  ungroup() |>
  gt() |>
  tab_header(title = md("**Year Over Year Increase in Types of Plastic Use**"),
             subtitle = "Calculated from 2020") |>
  fmt_percent(columns = `2019_2020_YoY`,
              decimals = 2) |>
  cols_label(plastic_type = "Type",
             `2019_2020_YoY` = "YoY Increase") |>
  tab_options(heading.background.color = "#EB6864",
              column_labels.background.color = "lightgrey",
              row.striping.background_color = "lightgrey",
              row.striping.include_table_body = TRUE) |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels(everything())) |>
  tab_style(style = cell_borders(sides = "bottom",
            color = "#333333",
            weight = px(2)),
            locations = list(cells_column_labels(), cells_title(groups = "subtitle"))) |>
  tab_options(
    data_row.padding = px(16),
    heading.padding = px(10)) |>
  data_color(columns = `2019_2020_YoY`,
             colors = col_numeric(
             palette = c("lightgreen", "white", "#D95C5C"),
             domain = NULL))
```

#### Interpretation

The year-over-year results reinforce the patterns seen in the animated and slope chart visuals. While those visuals show that polypropylene (PP) and several other plastics increase in proportion, the YoY table highlights how large those increases are in relative terms, with **PP** and **PVC** showing the most dramatic growth (+243.12% and +364.36% respectively). At the same time, the substantial decrease in the **“other”** category aligns with its visible decline in the earlier visuals (-48.88%).

Together, these results confirm that the shift from 2019 to 2020 is not only visible in composition but also reflected in large relative changes in specific plastic types, suggesting that a few categories are driving most of the change.

:::

### Column 2
#### Country Plot
:::{card}

```{r}
#| label: Figure 2
#| fig-alt: "Scatterplot comparing total plastic waste by country in 2019 and 2020, with a dashed line indicating no change between years"
#| fig-cap: "Figure 2 compares total plastic waste by country. Each point represents a country, and countries above the dashed line reported higher plastic waste in 2020 than in 2019, while those below the line reported decreases."

country_change <- plastics |>
  filter(parent_company != "Grand Total",
         country != "EMPTY") |>
  group_by(country, year) |>
  summarize(total_plastic = sum(grand_total, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = year, values_from = total_plastic) |>
  filter(!is.na(`2019`), !is.na(`2020`))

outliers <- country_change |>
  slice_max(`2020`, n = 2)

country_change |>
  ggplot(aes(x = `2019`, y = `2020`)) +
  geom_point(color = "#D95C5C", alpha = 0.7, size = 3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  geom_text_repel(data = outliers,
                  aes(label = country),
                  size = 4,
                  fontface = "bold") +
  labs(title = "Plastic Waste by Country Before and During COVID",
       subtitle = "Countries above the dashed line reported more plastic waste in 2020 than in 2019",
       x = "Total Plastic Waste in 2019",
       y = "Total Plastic Waste in 2020") +
  theme_light()
```

This scatter plot shows us that most countries experienced relatively small changes in plastic waste between 2019 and 2020, as indicated by the clustering of points near the dashed line. However, a few countries most notably the **Philippines** and **Switzerland** stand out with substantially larger increases. This suggests that while global plastic waste patterns did shift during COVID-19, much of the change may be driven by a small number of outlier countries.

:::

## Row 5
**Conclusion**

While the global analysis reveals clear shifts in plastic composition from 2019 to 2020, it does not explain whether these changes occurred consistently across all countries or were driven by specific regions. The country-level results suggest that a small number of countries may be responsible for much of the observed change.

To better understand how plastic usage evolved at a more localized level, we focus on a single country case study. By examining changes within one country, we can explore how plastic composition shifted in response to the COVID-19 pandemic in greater detail.


# China Case Study

## Row 1

::: {.card}
### Why China?

China was selected as a case of change because its plastic composition shifts substantially from 2019 to 2020, particularly with a large increase in the **“other”** plastic category and a decline in **PET**, which seems to differ greatly from what we observed in border study. These changes highlight how plastic waste composition can vary over time within a single country as illustrated by the animated bar graph below. Additionally, we were interested in investigating how the COVID-19 pandemic may have impacted China specifically and influenced patterns of plastic usage and waste.


```{r}
#| out-width: 100%

china_data <- plastics |>
  filter(
    country == "China",
    parent_company != "Grand Total"
  ) |>
  pivot_longer(
    cols = c(hdpe, ldpe, o, pet, pp, ps, pvc),
    names_to = "plastic_type",
    values_to = "plastic_amount"
  ) |>
  group_by(year, plastic_type) |>
  summarize(
    total = sum(plastic_amount, na.rm = TRUE),
    .groups = "drop"
  ) |>
  group_by(year) |>
  mutate(proportion = total / sum(total)) |>
  ungroup()

china_plot <- china_data |>
  ggplot(aes(
    x = plastic_type,
    y = proportion,
    fill = plastic_type
  )) +
  geom_col(show.legend = FALSE) +
  scale_y_continuous(labels = label_percent()) +
  scale_fill_manual(values = journal_colors) +
  labs(
    title = "China Plastic Composition in {closest_state}",
    x = "Plastic Type",
    y = "Proportion"
  ) +
  theme_light() +
  theme(
    plot.title = element_text(face = "bold", size = 16)
  ) +
  transition_states(
    states = year,
    transition_length = 2,
    state_length = 1
  ) +
  ease_aes("cubic-in-out")

animate(
  china_plot,
  nframes = 80,
  fps = 10)
```
:::
### Column 2

::: {.panel-tabset}
#### Data Background
The geographic boundaries used in this Province COVID-19 Cases map come from the Natural Earth dataset via the R package rnaturalearth, accessed through the rnaturalearth package in R. This dataset provides standardized, publicly available map data at the country and subnational (province) level. We used province-level boundaries for China to spatially visualize COVID-19 case data and examine how pandemic intensity varied across regions.

The COVID-19 data used in this analysis comes from the Johns Hopkins University Center for Systems Science and Engineering (JHU CSSE) COVID-19 Data Repository, which collected global case data from January 22, 2020 through March 10, 2023. This dataset aggregates information from multiple official sources and provides detailed case counts at both global and subnational levels. We use province-level data for China to examine how the intensity of COVID-19 varied geographically during 2020.

#### Province COVID-19 Cases
```{r}
# Data extracted from many places including the WHO, Collected by JHU
# https://github.com/CSSEGISandData/COVID-19/blob/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv
china_cases <- read.csv(here('time_series_covid19_confirmed_global.csv')) |>
  filter(Country.Region == 'China')

china_cases_clean <- china_cases |>
  select(-Lat, -Long, -Country.Region) |>
  pivot_longer(cols = -Province.State,
               names_to = "date",
               values_to = "cases") |>
  mutate(date = str_remove(date, "^X"),
         date = mdy(date)) |>
  arrange(Province.State, date) |>
  group_by(Province.State) |>
  mutate(new_cases = cases - lag(cases, default = 0)) |>
  ungroup() |>
  group_by(Province.State) |>
  summarize(total_new_cases = sum(new_cases, na.rm = TRUE),
            peak_cases = max(new_cases, na.rm = TRUE),
            .groups = "drop") |>
  rename(province = Province.State) |>
  arrange(desc(total_new_cases)) |>
  mutate(province = str_trim(province),
         province = case_when(province %in% c("Hong Kong", "Macau", "Unknown") ~ NA, TRUE ~ province),
         province = case_when(province == "Inner Mongolia" ~ "Inner Mongol",
                              province == "Tibet" ~ "Xizang",
                              TRUE ~ province)) |>
  filter(!is.na(province))
  
# From the rnaturalearth package  
china_map <- ne_states(country = "china", returnclass = "sf") |>
  select(name, geometry)

Coords_Cases_China <- china_map |>
  inner_join(china_cases_clean, by= join_by(name == province))

pal <- colorNumeric(palette = journal_colors,
                    domain = Coords_Cases_China$total_new_cases)

Coords_Cases_China |>
  leaflet() |>
  addTiles() |>
  addControl(html = "COVID-19 Cases by Province in China (2020)",
             position = "topright")  |>
  addPolygons(fillColor = ~pal(total_new_cases),
              color = "lightgrey",
              weight = 1,
              fillOpacity = 1,
              popup = ~glue("<strong>{name}</strong><br> Cases: {comma(total_new_cases)}"),
              label = ~glue("{name}: {comma(total_new_cases)}")) |>
  addLegend(pal = pal,
            values = ~total_new_cases,
            title = "COVID Cases",
            position = "bottomright")
```

#### How COVID Shaped Plastic Use
The Province COVID-19 Cases map shows that COVID-19 cases were concentrated in specific regions of China, particularly in **Hubei** and **Guangdong** provinces (72,131 and 103,248 cases respectively). These regions likely experienced stricter lockdown measures, increased reliance on delivery services, and greater use of personal protective equipment like masks.

These changes in daily behavior may have contributed to the shifts in plastic types used by China. Increased food delivery, online shopping, and medical supply consumption would lead to greater use of mixed or less easily categorized plastics, which may explain the large increase in the **“other”** category. At the same time, disruptions to manufacturing and recycling systems, as well as changes in consumer demand, may have reduced the relative use of more standardized plastics such as **PET**, which is commonly used in beverage packaging.

Together, the geographic concentration of COVID-19 cases attempts to explain why China’s plastic composition shifted differently from the global pattern, highlighting how localized pandemic impacts can influence material usage and waste generation.
:::

